Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2019-10-25 write-up/fda-recurrent.tex
Fixing up paper and the repo

\title[Recurrent event analysis with functional covariates via random
subsampling]{Recurrent event analysis in the presence of real-time
  high frequency data via random subsampling}
\author{Walter Dempsey}
\address {Department of Statistics, 
  Harvard University, One Oxford Street
   Cambridge, MA  02138, USA}
 \email{wdempsey@fas.harvard.edu}

@@

\begin{abstract}
Digital monitoring studies collect real-time high frequency data via
mobile sensors in the subjects' natural environment.  
This data can be used to model the impact of changes in physiology on
recurrent event outcomes such as smoking, drug use, alcohol use, or
self-identified moments of suicide ideation. 
Likelihood calculations for the recurrent event analysis, however,
become computationally prohibitive in this setting.  
% Unlike classic joint models, event risk is a function of
% recent sensor data, not simply current sensor values.  
Motivated by this, a random subsampling framework is proposed for
computationally efficient, approximate likelihood-based estimation.
A subsampling-unbiased estimator for the derivative of the cumulative
hazard enters into an approximation of log-likelihood. The estimator
has two sources of variation: the first due to the recurrent event
model and the second due to subsampling. The latter can be reduced by
increasing the sampling rate; however, this leads to increased
computational costs.  The approximate score equations are equivalent
to logistic regression score equations, allowing for standard,
``off-the-shelf'' software to be used in fitting these models.
Simulations demonstrate the method and efficiency-computation
tradeoff. We then illustrate the approach using data from a digital
monitoring study of suicide ideation.
\end{abstract}

@@

The goal of this paper is two-fold: (1) to discuss the appropriate
choice of statistical model for joint high-frequency sensor data and
the recurrent event data, and (2) to construct a robust,
easy-to-implement method for parameter estimation and inference. 
For (1), we discuss an important issue regarding measurement-error
models when paired with recurrent event outcomes.
For (2), we introduce a random subsampling procedure that has
several benefits.  First, the resulting inference is unbiased;
however, there is a computation-efficiency trade-off. In particular, a
higher sampling rate can decrease estimator variance at the cost of
increased computation.  We show via simulations that the benefits of
incredibly high sampling rates is often negligible, as the
contribution to the variation is small in relation to to variation in
the underlying stochastic processes. Second, derived estimating
equations are optimal, implying loss of statistical efficiency is only
due to subsampling procedure and not derived methodology.  Finally,
implementation can leverage existing, standard software for functional
data analysis and logistic regression, leading to fast adoption by
domain scientists.

>>>>

\title[Recurrent event analysis with functional covariates via random subsampling]{Recurrent event analysis in the presence of real-time high frequency data via random subsampling}
\author{Walter Dempsey}
\address {Department of Biostatistics, University of Michigan, 1415 Washington Heights, Ann Arbor, MI 48109, USA}
 \email{wdem@umich.edu}

@@

\begin{abstract}
Digital monitoring studies collect real-time high frequency data via mobile sensors in the subjects' natural environment. This data can be used to model the impact of changes in physiology on recurrent event outcomes such as smoking, drug use, alcohol use, or self-identified moments of suicide ideation. Likelihood calculations for the recurrent event analysis, however, become computationally prohibitive in this setting. Motivated by this, a random subsampling framework is proposed for computationally efficient, approximate likelihood-based estimation. A subsampling-unbiased estimator for the derivative of the cumulative hazard enters into an approximation of log-likelihood. The estimator has two sources of variation: the first due to the recurrent event model and the second due to subsampling. The latter can be reduced by increasing the sampling rate; however, this leads to increased computational costs.  The approximate score equations are equivalent to logistic regression score equations, allowing for standard,
``off-the-shelf'' software to be used in fitting these models. Simulations demonstrate the method and efficiency-computation trade-off. We then illustrate the approach using data from a digital monitoring study of suicidal ideation.
\end{abstract}

@@

The goal of this paper is two-fold: (1) to discuss the appropriate choice of statistical model for joint high-frequency sensor data and the recurrent event data, and (2) to construct a simple, easy-to-implement method for parameter estimation and inference. For (1), we discuss an important issue regarding measurement-error models when paired with recurrent event outcomes. For (2), we introduce a random subsampling procedure that has several benefits.  First, the resulting inference is unbiased; however, there is a computation-efficiency trade-off. In particular, a
higher sampling rate can decrease estimator variance at the cost of increased computation.  We show via simulations that the benefits of incredibly high sampling rates is often negligible, as the contribution to the variation is small in relation to to variation in the underlying stochastic processes. Second, derived estimating equations are optimal, implying loss of statistical efficiency is only due to subsampling procedure and not derived methodology.  Finally,
implementation can leverage existing, standard software for functional data analysis and logistic regression, leading to fast adoption by domain scientists.
